Tutorials: A Brief Overview

### 2023-04-30 temp.qmd
updates to manucript figs

---
title: "Tutorials: A Brief Overview"
format:
  html:
    page-layout: full
    grid:
      sidebar-width: 0px
      body-width: 2000px
      margin-width: 0px
      gutter-width: 1.5rem
    mermaid:
      theme: forest
#    css: styles.css
#    theme: cosmo
#    fontsize: 1.1em
#    linestretch: 1.7

@@

@@

>>>>

---
#title: "Tutorials: A Brief Overview"
format:
  html:
    page-layout: full
    grid:
      #sidebar-width: 250px
      sidebar-width: 0px
      body-width: 2000px
      margin-width: 0px
      #gutter-width: 1.5rem
    #mermaid:
      #theme: forest
    css: styles.css
#    theme: cosmo
#    fontsize: 1.1em
#    linestretch: 1.7

@@

<script src="https://cdn.tailwindcss.com"></script>

```{html}
<div class="alert alert-primary" role="alert">
  This is a primary alert—check it out!
</div>
```

xxxyyyzzz
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```       
\begin{tabular}[t]{@{}l}
This is where the reciepient goes \\ 
Address 1 \\
Address 2 \\
Address 3 \\
Address 3 \\
\end{tabular}
\hfill
\begin{tabular}[t]{@{}l}
This is where the sender goes \\  
Person \\
Department \\ 
Address 1 \\ 
Address 2 \\  
Address 3 \\  
Address 4 \\  
Address 5 \\  
Address 6 \\ 
Telephone \\  
Fax \\  
Email \\
\end{tabular}

xxxyyyzzz
xxxxx

<div class="alert alert-primary" role="alert">
  This is a primary alert with <a href="#" class="alert-link">an example link</a>. Give it a click if you like.
</div>

<span class="badge badge-pill badge-primary">Primary</span>

<a href="#" class="badge badge-primary">Primary</a>

<div class="card" style="width: 18rem;">
  <img class="card-img-top" src="..." alt="Card image cap">
  <div class="card-body">
    <div class="card-title">Card title</div>
    <p class="card-text">Some quick example text to build on the card title and make up the bulk of the card's content.</p>
    <a href="#" class="btn btn-primary">Go somewhere</a>
  </div>
</div>

<div class="card">
  <div class="card-body">
    This is some text within a card body.
  </div>
</div>

<div class="card" style="width: 18rem;">
  <img class="card-img-top" src="..." alt="Card image cap">
  <div class="card-body">
    <h5 class="card-title">Card title</h5>
    <p class="card-text">Some quick example text to build on the card title and make up the bulk of the card's content.</p>
  </div>
  <ul class="list-group list-group-flush">
    <li class="list-group-item">Cras justo odio</li>
    <li class="list-group-item">Dapibus ac facilisis in</li>
    <li class="list-group-item">Vestibulum at eros</li>
  </ul>
  <div class="card-body">
    <a href="#" class="card-link">Card link</a>
    <a href="#" class="card-link">Another link</a>
  </div>
</div>

<div class="card">
  <div class="card-header">
    Featured
  </div>
  <div class="card-body">
    <h5 class="card-title">Special title treatment</h5>
    <p class="card-text">With supporting text below as a natural lead-in to additional content.</p>
    <a href="#" class="btn btn-primary">Go somewhere</a>
  </div>
</div>

<div class="card">
  <h5 class="card-header">Featured</h5>
  <div class="card-body">
    <h5 class="card-title">Special title treatment</h5>
    <p class="card-text">With supporting text below as a natural lead-in to additional content.</p>
    <a href="#" class="btn btn-primary">Go somewhere</a>
  </div>
</div>

<div class="card text-white bg-primary mb-3" style="max-width: 18rem;">
  <div class="card-header">Header</div>
  <div class="card-body">
    <h5 class="card-title">Primary card title</h5>
    <p class="card-text">Some quick example text to build on the card title and make up the bulk of the card's content.</p>
  </div>
</div>

<div class="card border-primary mb-3" style="max-width: 18rem;">
  <div class="card-header">Header</div>
  <div class="card-body text-primary">
    <h5 class="card-title">Primary card title</h5>
    <p class="card-text">Some quick example text to build on the card title and make up the bulk of the card's content.</p>
  </div>
</div>

<div class="card border-success mb-3" style="max-width: 18rem;">
  <div class="card-header bg-transparent border-success">Header</div>
  <div class="card-body text-success">
    <h5 class="card-title">Success card title</h5>
    <p class="card-text">Some quick example text to build on the card title and make up the bulk of the card's content.</p>
  </div>
  <div class="card-footer bg-transparent border-success">Footer</div>
</div>
<div class="alert alert-success" role="alert">
  <h4 class="alert-heading">Well done!</h4>
  <p>Aww yeah, you successfully read this important alert message. This example text is going to run a bit longer so that you can see how spacing within an alert works with this kind of content.</p>
  <hr>
  <p class="mb-0">Whenever you need to, be sure to use margin utilities to keep things nice and tidy.</p>
</div>

<dl class="row">
  <dt class="col-sm-3">Description lists</dt>
  <dd class="col-sm-9">A description list is perfect for defining terms.</dd>

  <dt class="col-sm-3">Euismod</dt>
  <dd class="col-sm-9">
    <p>Vestibulum id ligula porta felis euismod semper eget lacinia odio sem nec elit.</p>
    <p>Donec id elit non mi porta gravida at eget metus.</p>
  </dd>

  <dt class="col-sm-3">Malesuada porta</dt>
  <dd class="col-sm-9">Etiam porta sem malesuada magna mollis euismod.</dd>

  <dt class="col-sm-3 text-truncate">Truncated term is truncated</dt>
  <dd class="col-sm-9">Fusce dapibus, tellus ac cursus commodo, tortor mauris condimentum nibh, ut fermentum massa justo sit amet risus.</dd>

  <dt class="col-sm-3">Nesting</dt>
  <dd class="col-sm-9">
    <dl class="row">
      <dt class="col-sm-4">Nested definition list</dt>
      <dd class="col-sm-8">Aenean posuere, tortor sed cursus feugiat, nunc augue blandit nunc.</dd>
    </dl>
  </dd>
</dl>

<!-- Button trigger modal -->
<button type="button" class="btn btn-primary" data-toggle="modal" data-target="#exampleModal">
  Launch demo modal
</button>

<!-- Modal -->
<div class="modal fade" id="exampleModal" tabindex="-1" role="dialog" aria-labelledby="exampleModalLabel" aria-hidden="true">
  <div class="modal-dialog" role="document">
    <div class="modal-content">
      <div class="modal-header">
        <h5 class="modal-title" id="exampleModalLabel">Modal title</h5>
        <button type="button" class="close" data-dismiss="modal" aria-label="Close">
          <span aria-hidden="true">&times;</span>
        </button>
      </div>
      <div class="modal-body">
        ...
      </div>
      <div class="modal-footer">
        <button type="button" class="btn btn-secondary" data-dismiss="modal">Close</button>
        <button type="button" class="btn btn-primary">Save changes</button>
      </div>
    </div>
  </div>
</div>

<button type="button" class="btn btn-lg btn-danger" data-toggle="popover" title="Popover title" data-content="And here's some amazing content. It's very engaging. Right?">Click to toggle popover</button>

<div class="w-25 p-3" style="background-color: #eee;">Width 25%</div>
<div class="w-50 p-3" style="background-color: #eee;">Width 50%</div>
<div class="w-75 p-3" style="background-color: #eee;">Width 75%</div>
<div class="w-100 p-3" style="background-color: #eee;">Width 100%</div>

<div style="height: 100px; background-color: rgba(255,0,0,0.1);">
  <div class="h-25 d-inline-block" style="width: 120px; background-color: rgba(0,0,255,.1)">Height 25%</div>
  <div class="h-50 d-inline-block" style="width: 120px; background-color: rgba(0,0,255,.1)">Height 50%</div>
  <div class="h-75 d-inline-block" style="width: 120px; background-color: rgba(0,0,255,.1)">Height 75%</div>
  <div class="h-100 d-inline-block" style="width: 120px; background-color: rgba(0,0,255,.1)">Height 100%</div>
</div>

<p class="text-justify">Ambitioni dedisse scripsisse iudicaretur. Cras mattis iudicium purus sit amet fermentum. Donec sed odio operae, eu vulputate felis rhoncus. Praeterea iter est quasdam res quas ex communi. At nos hinc posthac, sitientis piros Afros. Petierunt uti sibi concilium totius Galliae in diem certam indicere. Cras mattis iudicium purus sit amet fermentum.</p>

<p class="text-left">Left aligned text on all viewport sizes.</p>
<p class="text-center">Center aligned text on all viewport sizes.</p>
<p class="text-right">Right aligned text on all viewport sizes.</p>

<p class="text-sm-left">Left aligned text on viewports sized SM (small) or wider.</p>
<p class="text-md-left">Left aligned text on viewports sized MD (medium) or wider.</p>
<p class="text-lg-left">Left aligned text on viewports sized LG (large) or wider.</p>
<p class="text-xl-left">Left aligned text on viewports sized XL (extra-large) or wider.</p>

<p class="font-weight-bold">Bold text.</p>
<p class="font-weight-normal">Normal weight text.</p>
<p class="font-weight-light">Light weight text.</p>
<p class="font-italic">Italic text.</p>

<table class="table table-bordered">
  <thead>
    <tr>
      <th scope="col">#</th>
      <th scope="col">First</th>
      <th scope="col">Last</th>
      <th scope="col">Handle</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <th scope="row">1</th>
      <td>Mark</td>
      <td>Otto</td>
      <td>@mdo</td>
    </tr>
    <tr>
      <th scope="row">2</th>
      <td>Jacob</td>
      <td>Thornton</td>
      <td>@fat</td>
    </tr>
    <tr>
      <th scope="row">3</th>
      <td colspan="2">Larry the Bird</td>
      <td>@twitter</td>
    </tr>
  </tbody>
</table>

<blockquote class="blockquote">
  <p class="mb-0">Lorem ipsum dolor sit amet, consectetur adipiscing elit. Integer posuere erat a ante.</p>
</blockquote>

<table class="table">
  <thead>
    <tr>
      <th scope="col">#</th>
      <th scope="col">First</th>
      <th scope="col">Last</th>
      <th scope="col">Handle</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <th scope="row">1</th>
      <td>Mark</td>
      <td>Otto</td>
      <td>@mdo</td>
    </tr>
    <tr>
      <th scope="row">2</th>
      <td>Jacob</td>
      <td>Thornton</td>
      <td>@fat</td>
    </tr>
    <tr>
      <th scope="row">3</th>
      <td>Larry</td>
      <td>the Bird</td>
      <td>@twitter</td>
    </tr>
  </tbody>
</table>

<table class="hover:table-fixed">
  <thead>
    <tr>
      <th>Song</th>
      <th>Artist</th>
      <th>Year</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <td>The Sliding Mr. Bones (Next Stop, Pottersville)</td>
      <td>Malcolm Lockyer</td>
      <td>1961</td>
    </tr>
    <tr>
      <td>Witchy Woman</td>
      <td>The Eagles</td>
      <td>1972</td>
    </tr>
    <tr>
      <td>Shining Star</td>
      <td>Earth, Wind, and Fire</td>
      <td>1975</td>
    </tr>
  </tbody>
</table>

::: {.column-screen}
![-](./figures/Figlanding5.svg)
:::

file:///Users/shawes/git/swhawes.github.io/docs/manuscript.html
xxxxx

```{dot}
digraph "Linux_kernel_diagram" {
	fontname="Helvetica,Arial,sans-serif"
	node [fontname="Helvetica,Arial,sans-serif"]
	edge [fontname="Helvetica,Arial,sans-serif"]
	graph [
		newrank = true,
		nodesep = 0.3,
		ranksep = 0.2,
		overlap = true,
		splines = false,
	]
	node [
		fixedsize = false,
		fontsize = 24,
		height = 1,
		shape = box,
		style = "filled,setlinewidth(5)",
		width = 2.2
	]
	edge [
		arrowhead = none,
		arrowsize = 0.5,
		labelfontname = "Ubuntu",
		weight = 10,
		style = "filled,setlinewidth(5)"
	]
	subgraph system {
		node [color = "#e27dd6ff"]
		edge [color = "#e27dd6ff"]
		system_ [
			fixedsize = true,
			height = 0,
			shape = point,
			style = invis,
			shape = point
		]
		system [
			URL = "https://en.wikibooks.org/wiki/The_Linux_Kernel/System",
			fillcolor = white,
			fixedsize = true,
			height = 0.6,
			row = func,
			width = 2]
		system -> system_ [
			arrowhead = "",
			row = func];
		SCI [
			URL = "https://en.wikibooks.org/wiki/The_Linux_Kernel/Syscalls",
			fillcolor = "#d9e7ee",
			fixedsize = true,
			label = "System calls",
			row = usr,
			shape = ellipse]
		sysfs [
			fillcolor = "#b2d3e4",
			label = "proc & sysfs\nfile systems"]
		SCI -> sysfs
		DM [
			fillcolor = "#91b5c9",
			fixedsize = true,
			fontsize = 20,
			height = 0.8,
			label = "Device\nModel",
			shape = octagon,
			width = 2]
		sysfs -> DM
		log_sys [
			fillcolor = "#6a9ab1",
			fontsize = 20,
			label = "system run,\nmodules,\ngeneric\nHW access "]
		DM -> log_sys
		bus_drv [
			fillcolor = "#71809b",
			label = "bus drivers"]
		log_sys -> bus_drv
		buses [
			fillcolor = "#777777",
			fontcolor = white,
			fontsize = 20,
			label = "buses:\nPCI, USB ...",
			row = chip]
		bus_drv -> buses
	}
	subgraph networking {
		node [color = "#61c2c5"]
		edge [color = "#61c2c5"]
		networking_ [
			fixedsize = true,
			height = 0,
			shape = point,
			style = invis,
			shape = point
				width = 0]
		networking [
			URL = "https://en.wikibooks.org/wiki/The_Linux_Kernel/Networking",
			fillcolor = white,
			fixedsize = true,
			height = 0.6,
			row = func,
			width = 2]
		networking -> networking_ [
			arrowhead = "",
			row = func]
		sock [
			fillcolor = "#d9e7ee",
			fixedsize = true,
			label = Sockets,
			row = usr,
			shape = ellipse]
		prot_fam [
			fillcolor = "#b2d3e4",
			label = "protocol\nfamilies"]
		sock -> prot_fam
		log_prot [
			fillcolor = "#6a9ab1",
			label = "protocols:\nTCP, UDP, IP"]
		prot_fam -> log_prot
		netif [
			fillcolor = "#71809b",
			fontsize = 20,
			label = "network\ninterfaces\nand drivers"]
		log_prot -> netif
		net_hw [
			fillcolor = "#777777",
			fontcolor = white,
			fontsize = 20,
			label = "network:\nEthernet, WiFi ...",
			row = chip]
		netif -> net_hw
		NFS [
			color = "#8383cc",
			fillcolor = "#91b5c9",
			fixedsize = true,
			height = 0.8,
			label = NFS,
			shape = octagon,
			width = 1.2]
		NFS -> log_prot [weight = 0]
	}
	subgraph processing {
		node [color = "#c46747"]
		edge [color = "#c46747"]
		processing_ [
			fixedsize = true,
			height = 0,
			shape = point
				style = invis,
			width = 0]
		processing [
			URL = "https://en.wikibooks.org/wiki/The_Linux_Kernel/Processing",
			fillcolor = white,
			fixedsize = true,
			height = 0.6,
			row = func,
			width = 2]
		processing -> processing_ [
			arrowhead = "",
			row = func]
		proc [
			fillcolor = "#d9e7ee",
			fixedsize = true,
			label = Processes,
			row = usr,
			shape = ellipse]
		Tasks [
			fillcolor = "#b2d3e4"]
		proc -> Tasks
		sync [
			fillcolor = "#91b5c9",
			fixedsize = true,
			fontsize = 20,
			fontname = "Arial Narrow"
			label = synchronization,
			height = 0.7,
			//width = 2,
			shape = octagon]
		Tasks -> sync
		sched [
			fillcolor = "#6a9ab1",
			label = Scheduler]
		sync -> sched
		IRQ [
			fillcolor = "#71809b",
			fontsize = 20,
			label = "interrupts\ncore,\nCPU arch"]
		sched -> IRQ
		CPU [
			fillcolor = "#777777",
			fontcolor = white,
			fontsize = 20,
			row = chip]
		IRQ -> CPU
	}	// processing
	subgraph mem {
		node [
			color = "#51bf5b",
			height = 1
		]
		edge [color = "#51bf5b"]
		MA [
			color = "#51bf5b",
			fillcolor = "#d9e7ee",
			fixedsize = true,
			label = "memory\naccess",
			row = usr,
			height = 1,
			shape = ellipse]
		MA -> VM
		mmap [
			fillcolor = "#91b5c9",
			fixedsize = true,
			fontsize = 20,
			height = 0.8,
			label = "memory\nmapping",
			shape = octagon,
			width = 2]
		mmap -> log_mem
		log_mem -> PA
		SW [
			color = "#8383cc",
			fillcolor = "#91b5c9",
			fixedsize = true,
			label = Swap,
			height = 0.8,
			shape = octagon,
			width = 1.2]
		mmap -> SW [weight = 1]
		SW -> block [
			color = "#8383cc", weight = 1]
		PA [
			fillcolor = "#71809b",
			label = "Page\nAllocator"
		]
		PC -> PA [weight = 0 color="#51bf5b"]
		RAM [
			color = "#51bf5b",
			fillcolor = "#777777",
			fontcolor = white,
			fontsize = 20,
			label = "MMU, RAM",
			height = 1,
			row = chip]
		PA -> RAM
		memory -> memory_ [
			arrowhead = "",
			row = func]
		VM -> mmap
	}	// mem
	subgraph storage {
		node [color = "#8383cc"]
		edge [color = "#8383cc"]
		NFS;
		storage_ [
			shape = point,
			fixedsize = true,
			height = 0,
			style = invis,
			width = 0]
		storage [
			URL = "https://en.wikibooks.org/wiki/The_Linux_Kernel/Storage",
			fillcolor = white,
			fixedsize = true,
			height = 0.6,
			row = func,
			width = 2]
		storage -> storage_ [
			arrowhead = "",
			row = func]
		FS [
			fillcolor = "#d9e7ee",
			fixedsize = true,
			label = "files and\ndirectories",
			row = usr,
			shape = ellipse]
		VFS [
			fillcolor = "#b2d3e4",
			label = "Virtual\nFile System"]
		FS -> VFS
		VFS -> mmap [weight = 0]
		VFS -> NFS [weight = 0]
		logFS [
			fillcolor = "#6a9ab1",
			fontsize = 20,
			label = "logical\nfilesystems:\next3, xfs ..."]
		VFS -> logFS
		PC [
			fillcolor = "#91b5c9",
			fixedsize = true,
			fontsize = 20,
			height = 0.8,
			label = "page\ncache",
			shape = octagon,
			width = 1.2]
		VFS -> PC [weight = 0]
		block [
			fillcolor = "#71809b",
			fontsize = 20,
			label = "Block\ndevices\nand drivers"]
		logFS -> block
		SD [
			fillcolor = "#777777",
			fontcolor = white,
			fontsize = 20,
			label = "storage devices:\nSCSI, NVMe ...",
			row = chip]
		block -> SD
	}	// storge
	subgraph HI {
		node [color = "#cfbf57ff"]
		edge [
			color = "#cfbf57ff",
			weight = 10
		]
		HI_ [
			fixedsize = true,
			height = 0,
			shape = point,
			style = invis,
			width = 0]
		HI [
			URL = "https://en.wikibooks.org/wiki/The_Linux_Kernel/Human_interfaces",
			fillcolor = white,
			fixedsize = true,
			fontsize = 12,
			height = 0.6,
			label = "Forms\nof\nchange",
			row = func,
			width = 2]
		HI -> HI_ [
			arrowhead = "",
			row = func]
		char [
			fillcolor = "#d9e7ee",
			fixedsize = true,
			label = "char\ndevices",
			row = usr,
			shape = ellipse]
		input [
			fillcolor = "#b2d3e4",
			label = "input\nsubsystem"]
		char -> input
		F7 [
			fillcolor = "#6a9ab1",
			label = "HI class\ndrivers"]
		input -> F7
		HID [
			fillcolor = "#71809b",
			fontsize = 20,
			URL = "https://www.kernel.org/doc/html/latest/hid/",
			label = "HI\nperipherals\ndrivers"]
		F7 -> HID
		display [
			fillcolor = "#777777",
			fontcolor = white,
			fontsize = 19,
			label = "keyboard, mouse,\ndisplay, audio",
			row = chip]
		HID -> display
	} // HI
	subgraph functions {
		graph [rank = same]
		edge [
			style = invis,
			weight = 1
		]
		system;
		networking;
		system -> processing [weight = 1]
		storage -> networking [weight = 1]
		memory [
			color = "#51bf5b",
			URL = "https://en.wikibooks.org/wiki/The_Linux_Kernel/Memory",
			fillcolor = white,
			fixedsize = true,
			height = 0.6,
			row = func,
			width = 2]
		memory -> storage [weight = 1]
		processing -> memory [weight = 1]
		functions_ [
			fixedsize = true,
			height = 0,
			shape = point
			style = invis,
			width = 0]
		functions_ -> HI -> system [weight = 1]
		functions [
			color = gray,
			tooltip = "Columns represent main functionalities of the kernel",
			URL = "http://www.makelinux.net/ldd3/chp-1-sect-2.shtml",
			fillcolor = gray,
			fixedsize = true,
			height = 0.6,
			row = func,
			style = dashed,
			width = 1.6]
		functions -> functions_ [
			arrowhead = "",
			color = gray,
			style = "",
			weight = ""]
	}
	subgraph interfaces {
		graph [rank = same]
		SCI;
		sock;
		FS;
		proc;
		char;
		usr_ [
			fixedsize = true,
			height = 0,
			shape = point
				style = invis,
			width = 0.5]
		usr [
			fillcolor = "#d9e7eeff",
			fixedsize = true,
			label = "user space\ninterfaces",
			row = usr,
			shape = ellipse,
			style = "filled,setlinewidth(0)"]
		MA;
	}
	{
		edge [style = invis weight = 10 ]
		system_;
		SCI;
		system_ -> SCI;
		networking_;
		sock;
		networking_ -> sock;
		storage_;
		FS;
		storage_ -> FS;
		processing_;
		proc;
		processing_ -> proc;
		HI_;
		char;
		HI_ -> char;
		MA;
		memory_ [
			fixedsize = true,
			height = 0,
			shape = point,
			style = invis,
			width = 0]
		memory_ -> MA;
	}
	subgraph virtual {
		graph [rank = same]
		sysfs;
		prot_fam;
		VFS;
		Tasks;
		input;
		D0 [
			fixedsize = true,
			height = 0,
			shape = point,
			style = invis,
			width = 0]
		virt [
			fillcolor = "#b2d3e4",
			label = "virtual\nsubsystems",
			URL = "https://en.wikipedia.org/wiki/Proxy_pattern",
			tooltip = "proxy between standard user space interfaces and internal implementations",
			style = "filled,setlinewidth(0)"]
		VM [
			color = "#51bf5b",
			fillcolor = "#b2d3e4",
			label = "Virtual\nmemory"]
	}
	subgraph bridges {
		graph [rank = same]
		bridges [
			fillcolor = "#91b5c9",
			shape = octagon,
			tooltip = "bridges between uniform virtual interfaces and various implementations",
			URL = "https://en.wikipedia.org/wiki/Bridge_pattern",
			style = "filled,setlinewidth(0)"]
		DM;
		NFS;
		mmap;
		sync;
		E0 [
			fixedsize = true,
			height = 0,
			shape = point,
			style = invis,
			width = 0]
		//PC
	}
	subgraph logical {
		graph [rank = same]
		log_sys;
		log_prot;
		logFS;
		sched;
		F7;
		F0 [
			fixedsize = true,
			height = 0,
			shape = point,
			style = invis,
			width = 0]
		logical [
			fillcolor = "#6a9ab1",
			style = "filled,setlinewidth(0)"]
		log_mem [
			color = "#51bf5b",
			fillcolor = "#6a9ab1",
			label = "logical\nmemory"]
		//SW
	}
	subgraph HWI {
		graph [rank = same]
		HWI [
			fillcolor = "#71809b",
			label = "hardware\ninterfaces",
			style = "filled,setlinewidth(0)"]
		bus_drv;
		netif;
		block;
		//PA;
		IRQ;
		HID;
		G0 [
			fixedsize = true,
			height = 0,
			shape = point,
			style = invis,
			width = 0]
	}
	subgraph HW {
		graph [rank = same]
		HW [
			fillcolor = "#777777",
			fontcolor = white,
			label = "electronics,\nhardware",
			row = chip,
			style = "filled,setlinewidth(0)"]
		buses;
		net_hw;
		SD;
		CPU;
		display;
		H0 [
			fixedsize = true,
			height = 0,
			shape = point,
			style = invis,
			width = 0]
		RAM;
	}
	bottom [
		label = "© 2007-2022 Costa Shulyupin http://www.MakeLinux.net/kernel/diagram",
		URL = "http://www.MakeLinux.net/kernel/diagram",
		shape = plaintext,
		style = ""]
	CPU -> bottom [style = invis]
	layers [
		fillcolor = lightgray,
		tooltip = "Functionalities are divided to common layers. It is approximate division.",
		height = 0.1,
		style = "filled,setlinewidth(0)",
		width = 0.5]
	functions -> layers [style = invis ]
	usr -> usr_ [
		arrowhead = "",
		color = "#d9e7eeff",
		minlen = 2]
	usr -> virt [
		color = "#d9e7eeff"]
	virt -> D0 [
		arrowhead = "",
		color = "#b2d3e4",
		minlen = 2]
	virt -> bridges [
		color = "#b2d3e4"]
	bridges -> E0 [
		arrowhead = "",
		color = "#91b5c9",
		minlen = 2,
		style = "filled,setlinewidth(6)",
		weight = ""]
	bridges -> logical [
		color = "#91b5c9",
		style = "filled,setlinewidth(6)"]
	logical -> F0 [
		arrowhead = "",
		color = "#6a9ab1",
		minlen = 2,
		row = logical,
		style = "filled,setlinewidth(6)",
		weight = ""]
	logical -> HWI [
		color = "#6a9ab1",
		row = logical,
		style = "filled,setlinewidth(6)"]
	HWI -> G0 [
		arrowhead = "",
		color = "#71809b",
		minlen = 2,
		row = HWI,
		style = "filled,setlinewidth(6)",
		weight = ""]
	HWI -> HW [
		color = "#71809b",
		row = HWI,
		style = "filled,setlinewidth(6)"]
	HW -> H0 [
		arrowhead = "",
		color = "#777777",
		minlen = 2,
		row = chip,
		style = "filled,setlinewidth(6)",
		weight = ""]
	layers -> usr [
		arrowhead = "",
		color = gray,
		style = "filled,setlinewidth(1)"]
	LKD [
		fontsize = 40,
		label = "Linux kernel diagram",
		shape = plain,
		style = ""]
	LKD -> processing [style = invis]
}

@@

```{dot}
digraph "Linux_kernel_diagram" {
	fontname="Helvetica,Arial,sans-serif"
	node [fontname="Helvetica,Arial,sans-serif"]
	edge [fontname="Helvetica,Arial,sans-serif"]
	graph [
		newrank = true,
		nodesep = 0.3,
		ranksep = 0.2,
		overlap = true,
		splines = false,
	]
	node [
		fixedsize = false,
		fontsize = 24,
		height = 1,
		shape = box,
		style = "filled,setlinewidth(5)",
		width = 2.2
	]
	edge [
		arrowhead = none,
		arrowsize = 0.5,
		labelfontname = "Ubuntu",
		weight = 10,
		style = "filled,setlinewidth(5)"
	]
	subgraph system {
		node [color = "#e27dd6ff"]
		edge [color = "#e27dd6ff"]
		system_ [
			fixedsize = true,
			height = 0,
			shape = point,
			style = invis,
			shape = point
		]
		system [
			URL = "https://en.wikibooks.org/wiki/The_Linux_Kernel/System",
			fillcolor = white,
			fixedsize = true,
			height = 0.6,
			row = func,
			width = 2]
		system -> system_ [
			arrowhead = "",
			row = func];
		SCI [
			URL = "https://en.wikibooks.org/wiki/The_Linux_Kernel/Syscalls",
			fillcolor = "#d9e7ee",
			fixedsize = true,
			label = "System calls",
			row = usr,
			shape = ellipse]
		sysfs [
			fillcolor = "#b2d3e4",
			label = "proc & sysfs\nfile systems"]
		SCI -> sysfs
		DM [
			fillcolor = "#91b5c9",
			fixedsize = true,
			fontsize = 20,
			height = 0.8,
			label = "Device\nModel",
			shape = octagon,
			width = 2]
		sysfs -> DM
		log_sys [
			fillcolor = "#6a9ab1",
			fontsize = 20,
			label = "system run,\nmodules,\ngeneric\nHW access "]
		DM -> log_sys
		bus_drv [
			fillcolor = "#71809b",
			label = "bus drivers"]
		log_sys -> bus_drv
		buses [
			fillcolor = "#777777",
			fontcolor = white,
			fontsize = 20,
			label = "buses:\nPCI, USB ...",
			row = chip]
		bus_drv -> buses
	}
	subgraph networking {
		node [color = "#61c2c5"]
		edge [color = "#61c2c5"]
		networking_ [
			fixedsize = true,
			height = 0,
			shape = point,
			style = invis,
			shape = point
				width = 0]
		networking [
			URL = "https://en.wikibooks.org/wiki/The_Linux_Kernel/Networking",
			fillcolor = white,
			fixedsize = true,
			height = 0.6,
			row = func,
			width = 2]
		networking -> networking_ [
			arrowhead = "",
			row = func]
		sock [
			fillcolor = "#d9e7ee",
			fixedsize = true,
			label = Sockets,
			row = usr,
			shape = ellipse]
		prot_fam [
			fillcolor = "#b2d3e4",
			label = "protocol\nfamilies"]
		sock -> prot_fam
		log_prot [
			fillcolor = "#6a9ab1",
			label = "protocols:\nTCP, UDP, IP"]
		prot_fam -> log_prot
		netif [
			fillcolor = "#71809b",
			fontsize = 20,
			label = "network\ninterfaces\nand drivers"]
		log_prot -> netif
		net_hw [
			fillcolor = "#777777",
			fontcolor = white,
			fontsize = 20,
			label = "network:\nEthernet, WiFi ...",
			row = chip]
		netif -> net_hw
		NFS [
			color = "#8383cc",
			fillcolor = "#91b5c9",
			fixedsize = true,
			height = 0.8,
			label = NFS,
			shape = octagon,
			width = 1.2]
		NFS -> log_prot [weight = 0]
	}
	subgraph processing {
		node [color = "#c46747"]
		edge [color = "#c46747"]
		processing_ [
			fixedsize = true,
			height = 0,
			shape = point
				style = invis,
			width = 0]
		processing [
			URL = "https://en.wikibooks.org/wiki/The_Linux_Kernel/Processing",
			fillcolor = white,
			fixedsize = true,
			height = 0.6,
			row = func,
			width = 2]
		processing -> processing_ [
			arrowhead = "",
			row = func]
		proc [
			fillcolor = "#d9e7ee",
			fixedsize = true,
			label = Processes,
			row = usr,
			shape = ellipse]
		Tasks [
			fillcolor = "#b2d3e4"]
		proc -> Tasks
		sync [
			fillcolor = "#91b5c9",
			fixedsize = true,
			fontsize = 20,
			fontname = "Arial Narrow"
			label = synchronization,
			height = 0.7,
			//width = 2,
			shape = octagon]
		Tasks -> sync
		sched [
			fillcolor = "#6a9ab1",
			label = Scheduler]
		sync -> sched
		IRQ [
			fillcolor = "#71809b",
			fontsize = 20,
			label = "interrupts\ncore,\nCPU arch"]
		sched -> IRQ
		CPU [
			fillcolor = "#777777",
			fontcolor = white,
			fontsize = 20,
			row = chip]
		IRQ -> CPU
	}	// processing
	subgraph mem {
		node [
			color = "#51bf5b",
			height = 1
		]
		edge [color = "#51bf5b"]
		MA [
			color = "#51bf5b",
			fillcolor = "#d9e7ee",
			fixedsize = true,
			label = "memory\naccess",
			row = usr,
			height = 1,
			shape = ellipse]
		MA -> VM
		mmap [
			fillcolor = "#91b5c9",
			fixedsize = true,
			fontsize = 20,
			height = 0.8,
			label = "memory\nmapping",
			shape = octagon,
			width = 2]
		mmap -> log_mem
		log_mem -> PA
		SW [
			color = "#8383cc",
			fillcolor = "#91b5c9",
			fixedsize = true,
			label = Swap,
			height = 0.8,
			shape = octagon,
			width = 1.2]
		mmap -> SW [weight = 1]
		SW -> block [
			color = "#8383cc", weight = 1]
		PA [
			fillcolor = "#71809b",
			label = "Page\nAllocator"
		]
		PC -> PA [weight = 0 color="#51bf5b"]
		RAM [
			color = "#51bf5b",
			fillcolor = "#777777",
			fontcolor = white,
			fontsize = 20,
			label = "MMU, RAM",
			height = 1,
			row = chip]
		PA -> RAM
		memory -> memory_ [
			arrowhead = "",
			row = func]
		VM -> mmap
	}	// mem
	subgraph storage {
		node [color = "#8383cc"]
		edge [color = "#8383cc"]
		NFS;
		storage_ [
			shape = point,
			fixedsize = true,
			height = 0,
			style = invis,
			width = 0]
		storage [
			URL = "https://en.wikibooks.org/wiki/The_Linux_Kernel/Storage",
			fillcolor = white,
			fixedsize = true,
			height = 0.6,
			row = func,
			width = 2]
		storage -> storage_ [
			arrowhead = "",
			row = func]
		FS [
			fillcolor = "#d9e7ee",
			fixedsize = true,
			label = "files and\ndirectories",
			row = usr,
			shape = ellipse]
		VFS [
			fillcolor = "#b2d3e4",
			label = "Virtual\nFile System"]
		FS -> VFS
		VFS -> mmap [weight = 0]
		VFS -> NFS [weight = 0]
		logFS [
			fillcolor = "#6a9ab1",
			fontsize = 20,
			label = "logical\nfilesystems:\next3, xfs ..."]
		VFS -> logFS
		PC [
			fillcolor = "#91b5c9",
			fixedsize = true,
			fontsize = 20,
			height = 0.8,
			label = "page\ncache",
			shape = octagon,
			width = 1.2]
		VFS -> PC [weight = 0]
		block [
			fillcolor = "#71809b",
			fontsize = 20,
			label = "Block\ndevices\nand drivers"]
		logFS -> block
		SD [
			fillcolor = "#777777",
			fontcolor = white,
			fontsize = 20,
			label = "storage devices:\nSCSI, NVMe ...",
			row = chip]
		block -> SD
	}	// storge
	subgraph HI {
		node [color = "#cfbf57ff"]
		edge [
			color = "#cfbf57ff",
			weight = 10
		]
		HI_ [
			fixedsize = true,
			height = 0,
			shape = point,
			style = invis,
			width = 0]
		HI [
			URL = "https://en.wikibooks.org/wiki/The_Linux_Kernel/Human_interfaces",
			fillcolor = white,
			fixedsize = true,
			fontsize = 12,
			height = 0.6,
			label = "human\ninterface",
			row = func,
			width = 2]
		HI -> HI_ [
			arrowhead = "",
			row = func]
		char [
			fillcolor = "#d9e7ee",
			fixedsize = true,
			label = "char\ndevices",
			row = usr,
			shape = ellipse]
		input [
			fillcolor = "#b2d3e4",
			label = "input\nsubsystem"]
		char -> input
		F7 [
			fillcolor = "#6a9ab1",
			label = "HI class\ndrivers"]
		input -> F7
		HID [
			fillcolor = "#71809b",
			fontsize = 20,
			URL = "https://www.kernel.org/doc/html/latest/hid/",
			label = "HI\nperipherals\ndrivers"]
		F7 -> HID
		display [
			fillcolor = "#777777",
			fontcolor = white,
			fontsize = 19,
